---
title: "Coursera Machine Learning Project"
author: "Xenia Sokolova"
date: '2020, april 2 '
output: html_document
---




## COURSERA MACHINE LEARNING PROJECT 
## Weight Lifting Exercises Dataset

## EXECUTIVE SUMMARY 

We propose a dataset with 5 classes (sitting-down, standing-up, standing, walking, and sitting) collected on 8 hours of activities of 4 healthy subjects. The goal of the project is to predict the manner in which they did the exercise. 

## BASIC EXPLORATORY DATA ANALYSIS

Downloading r packages:
```{r, cache=TRUE, echo=TRUE}
library(readr)
library(caret)
library(randomForest)
library(dplyr)
```

Loading data:
```{r, cache=TRUE, echo=TRUE}
setwd("C:/Users/xssok/Documents")
download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", 
              "pml-training.csv")
download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", 
              "pml-testing.csv")

training_pml <-read.csv("pml-training.csv")
testing_pml <- read.csv("pml-testing.csv")
```

Checking dimensions of the training and testing data:
```{r, cache=TRUE, echo=TRUE}
dim(training_pml)
dim(testing_pml)
```

Cross validation:
```{r, cache=TRUE, echo=TRUE}
set.seed(1234)
inTrain <- createDataPartition(y=training_pml$classe,
                               p=0.6, list=FALSE)
training <- training_pml[inTrain,] ## subset data into training set
testing <- training_pml[-inTrain,] ## subset data into test set
```

Explore the training set:
```{r, cache=TRUE, echo=TRUE}
dim(training)

dim(testing)

str(training)

table(training$classe)
```

## FEATURE SELECTION

Exclude first 7 columns: "X", "user_name", "raw_timestamp_part_1", "raw_timestamp_part_2", "cvtd_timestamp", "new_window", "num_window", because we don't need them for the prediction.

```{r, cache=TRUE, echo=TRUE}
training_ex <- select(training, -(X:num_window))
```

Select variables with missing data (more than 95%) and exclude them from the data table:

```{r, cache=TRUE, echo=TRUE}
training_ex[training_ex==""] <- NA
NArate <- apply(training_ex, 2, function(x) sum(is.na(x)))/nrow(training_ex)
training_cl <- training_ex[!(NArate>0.95)]
```

Now we've got 53 variables:

```{r, cache=TRUE, echo=TRUE}
str(training_cl)
```

Applying PCA Since the number of variables are over 50:

```{r, cache=TRUE, echo=TRUE}
preProc <- preProcess(training_cl[,-53],method="pca",thresh=.8) #12 components to capture 80 percent of the variance 
preProc
preProc <- preProcess(training_cl[,-53],method="pca",thresh=.9) #18 components to capture 90 percent of the variance
preProc
preProc <- preProcess(training_cl[,-53],method="pca",thresh=.95) #24 components to capture 95 percent of the variance
preProc
preProc <- preProcess(training_cl[,-53],method="pca",pcaComp=24) 
training_pca <- predict(preProc,training_cl[,-53])
```

## APPLYING A MODEL

Our data have got non-bionominal outcome and large sample size, that's why we're using random forest method.

```{r, cache=TRUE, echo=TRUE}
modFit <- randomForest(training_cl$classe ~ .,   data=training_pca, do.trace=F)
print(modFit) # view results
```

Checking on the test data:

```{r, cache=TRUE, echo=TRUE}
testing_cl <- select(testing, -(X:num_window))
testing_cl[testing_cl==""] <- NA
NArate <- apply(testing_cl, 2, function(x) sum(is.na(x)))/nrow(testing_cl)
testing_cl <- testing_cl[!(NArate>0.95)]
testing_pca <- predict(preProc,testing_cl[,-53])
confusionMatrix(testing_cl$classe,predict(modFit,testing_pca))
```

## PREDICTING CLASSES FOR TEST DATA (SIZE OF 20)

```{r, cache=TRUE, echo=TRUE}
testing_pml_cl <- select(testing_pml, -(X:num_window))
testing_pml_cl[testing_pml_cl==""] <- NA
NArate <- apply(testing_pml_cl, 2, function(x) sum(is.na(x)))/nrow(testing_pml_cl)
testing_pml_cl <- testing_pml_cl[!(NArate>0.95)]
testing_pml_pca <- predict(preProc,testing_pml_cl[,-53])
testing_pml_cl$classe <- predict(modFit,testing_pml_pca)
cbind (as.character(testing_pml_cl$classe))
```

## CONCLUSION

In this analyses, 19622 observations from weight lifting exercise were used to analyze and predict correct body movement from others during the exercise. 60% of the total observations (11776 observations) were used to build a model by random forest method, and the rest of 40% of the observations (7846 observations) were used for model validation (cross-validation). 
The model statistics showed that the built model had the overall accuracy of 97% for the testing set, which is not overlapping with observations used to built the model. The sensitivity was in between 92%-99% and the specificity was over 99% for all classes.
Overall, the model is well developed to predict the exercise classes during weight lifting. Therefore, under those condition, the model is expected to perform over 95% accuracy.